Continuous Deployment and Schema Evolution \\in SQL Databases

### 2015-01-29 paper.tex
Small textual changes to sections I-III.

SQL databases are an integral part of many (web) applications. They provide strong guarantees about the storage and retrieval of persistent data which makes them useful for various software applications. However the environment of software and release engineers is drastically changing: shorter development iterations, more frequent deployments to the production environment, or even fully automated deployments to the production environments several times a day. For release engineers this brings additional challenges. In particular when evolving the database schema (tables, columns, foreign keys, constraints, etc) of the SQL database, or rolling back such changes without loss of data.

We can distill these challenges into two separate problems:
\begin{itemize}
  \item{Given the deployment of a new version of a (web) application, depending on the deployment strategy it might be required that both the old and the new version of the application run in parallel (\textit{Rolling Upgrades}, or \textit{Canary Releases}~\cite{Humble:2010:CDR:1869904}). This means that the schema on the SQL database needs to be compatible with the schema expected by both the old and the new version of the application.}
  \item{Operations which modify the database schema in a SQL database (\textit{DDL statements}) can be either non-blocking or blocking in nature, depending on the implementation of the operations in that specific SQL database. In case of non-blocking \textit{DDL statements}, SQL queries issued by the application may proceed, whereas with blocking \textit{DDL statements} these queries may be put on hold for the duration of the operation. This disruption can range from milliseconds to hours depending on the schema operation, the size of the affected table(s), and database implementation. One such example would be adding a new non-nullable column to an existing table in a PostgreSQL database. This operation is blocking in nature, preventing other queries from running while it is being executed. When applied to a table of 50 million records (see Section~\ref{sec:results}), the queries from the application can be blocked for 2.5 minutes.}
\end{itemize}

When service windows are short or even non-existent, these two problems make releasing new versions of applications frequently, problematic for release engineers. To address this problem we aim to find or implement a tool which would allow its users to not only alter their database schema in a non-blocking fashion, but also do this without affecting older and still running applications.

@@

We first examined existing approaches, tools, and research which aim to solve these problems. A good starting point is the seminal book ``Continuous Delivery''~\cite{Humble:2010:CDR:1869904}. This book describes three high-level approaches to deal with migrating data without loss of data or requiring downtime:

\begin{itemize}
  \item{\textbf{Approach 1}: Copy the database or schema and start recording transactions. When the copying has completed, apply the schema changes to the copied database or schema. Finally replay all the recorded database transactions on the copied database or schema to reach an equivalent state. Once this state has been achieved, it must either be maintained by continuing to record and playback transactions, or make the switch from the old to the new database or schema.}
  \item{\textbf{Approach 2}: When using \textit{Blue-Green deployments}, two instances of the database are available. Take the inactive database and apply the schema operations to it. In the mean time these two databases need to be kept synchronized. This can be done by using approach 1, or somehow ensure that the table under change remains in a read-only state until the active and inactive databases have switched roles.}
  \item{\textbf{Approach 3}: The last is an approach where schema operations are performed separately from your application deployments using operations which are non-blocking. As a consequence the application must be able to deal with either the old or the new database schema being in use, and construct SQL queries which can operate correctly on both schemas.}
\end{itemize}

@@

\subsubsection{Percona Toolkit~\cite{PerconaToolkit} \& Openark Kit~\cite{OpenarkKit}}
are tools designed to help DBAs perform administrative tasks on MySQL databases. \textit{oak-online-alter-table} and \textit{pt-online-schema-change} respectively are commands which create an empty copy of the a specified table (a ghost table) and apply schema operations on ghost table. These commands then proceed to install triggers on both tables, which insert, update, or delete the corresponding record in the other table when a record in their own table is inserted, updated, or deleted. A background task iterates over all records in the original table (in batches) and lazily copies records from the original table to the ghost table. When all records have been copied the original table and ghost table switch names atomically, ensuring that clients now use the ghost table when querying the original table name.

\subsubsection{TableMigrator~\cite{TableMigrator}}
is a tool which works very similar to Openark Kit and Percona Toolkit. But instead of using triggers to synchronize the two tables with each other,  TableMigrator requires tables to have an additional column which stores the date and time that record was last modified. Using this information corresponding records in both tables are synchronized with each other by passing over all records in multiple passes. Once a pass can be done relatively quickly, TableMigrator acquires an exclusive write lock on the original table --- blocking the entire table --- and proceeds to do a final pass. Once this has been done, it atomically swaps the names of the original and ghost table before releasing the write lock. Since this final pass can be done quickly, any application executing queries on that table is only very briefly blocked.

\subsubsection{SoundCloud's Large Hadron Migrator~\cite{SoundcloudLHM} \& Facebook's Online Schema Change~\cite{FacebookOSC}}
are both tools which closely follow approach 1. These tools create a copy of a table (including its contents) and record modifications to records in the original table using triggers into a special ``deltas'' table. Once they have copied the original table and its contents, they apply the specified schema operations to the ghost table before proceeding to replay the recorded modifications from the ``deltas'' table onto the ghost table. When the ``deltas'' table is almost empty, these tools acquire an exclusive write lock on the original table --- like \textit{TableMigrator} does --- and apply the last of the recorded modifications before atomically switch the names of both tables.

@@

\subsubsection{Imago}
Imago~\cite{Dumitras:2009:WUF:1656980.1657005} is a tool which follows approach 2. With Imago there are two production environments, called ``universes'' by the authors. Imago integrates with the application and its environment at both the ingress point (load balancer) to be able to switch users from one ``universe'' to the other, and at the egress point where the (web) application sends queries to the SQL database. When a new version of the application (requiring a change to the database schema) needs to be deployed, Imago bootstraps the inactive production environment with a copy of the active database, applies the required schema operations and then copies over the data from the old universe to the new universe. When this process is almost complete, it enforces a short period of read-only access to the SQL database, while it is switching users from the old ``universe'' to the new ``universe'' using the load balancer. Imago makes no claims in regards to its support for foreign keys.

@@

\begin{itemize}
  \item{None of the tools used in practice support foreign keys and primarily focus on MySQL databases.}
  \item{These tools are very limited in the sense that they often only support making small changes to the existing database schema. This forces software and release engineers to often release intermediate versions of the application, as the schema is altered in several steps.}
\end{itemize}

\section{Our approach} %(1.0)

@@

The previously discussed tools from practice all use approach 1 or a variation of it. They prove that this is a viable way for simple evolutions of the database schema. However where these tools all create a single ghost table, we aim to create one ghost table for each table that is under change, and ghost tables for all tables which in turn are dependent on these tables under change. For instance adding a new non-nullable column ``summary'' to the ``books'' table (see Figure~\ref{fig:migration-stages}), means we should also create a ghost table for the ``publications`` table since there is a foreign key pointing from the ``publications'' table to the ``books'' table. The other tables do not require ghost tables, as they do not have any foreign keys referencing either the ``books'' or ``publications'' table. This approach ensures that we can support foreign keys.

>>>>

SQL databases are an integral part of many (web) applications. They provide strong guarantees about the storage and retrieval of persistent data which makes them useful for various software applications.

For release engineers, the persistence of the data and especially the structure of this data as embedded in the application's SQL schema is at odds with today's need for continuous deployment. Challenges include combining zero downtime constraints with evolving database schemas (affecting tables, columns, foreign keys, constraints, etc.), and the need to be able to rollback changes without loss of data.

We can distill the challenge of continuous delivery in the presence of schema evolution into two separate problems:

First, given the deployment of a new version of a (web) application, depending on the deployment strategy it might be required that both the old and the new version of the application run in parallel (\textit{Rolling Upgrades}, or \textit{Canary Releases}~\cite{Humble:2010:CDR:1869904}). This means that the schema on the SQL database needs to be compatible with the schema expected by both the old and the new version of the application.

Second, operations which modify the database schema in a SQL database (\textit{DDL statements}) can be either non-blocking or blocking in nature, depending on the implementation of the operations in that specific SQL database. In case of non-blocking \textit{DDL statements}, SQL queries issued by the application may proceed, whereas with blocking \textit{DDL statements} these queries may be put on hold for the duration of the operation. This disruption can range from milliseconds to hours depending on the schema operation, the size of the affected table(s), and the database implementation. One such example would be adding a new non-nullable column to an existing table in a PostgreSQL database. This operation is blocking in nature, preventing other queries from running while it is being executed. When applied to a table of 50 million records (see Section~\ref{sec:results}), the queries from the application can be blocked for 2.5 minutes.

When service windows are short or even non-existent, these two problems make releasing new versions of applications frequently problematic for release engineers. To address this problem we aim to find or implement a tool which  allows its users to not only alter their database schema in a non-blocking fashion, but also do this without affecting older and still running applications.

@@

\subsection{Continuous Delivery}\label{sec:cd}

We first examine existing approaches, tools, and research aiming to solve these problems. Our starting point is the seminal book ``Continuous Delivery''~\cite{Humble:2010:CDR:1869904}. This book describes three high-level approaches to deal with migrating data without loss of data or requiring downtime:

\begin{itemize}
  \item{\textbf{Approach 1}: Copy the database or schema and start recording transactions. When the copying has completed, apply the schema changes to the copied database or schema. Finally replay all the recorded database transactions on the copied database or schema to reach an equivalent state. Once this state has been achieved, it must either be maintained by continuing to record and playback transactions, or make the switch from the old to the new database or schema.}
  \item{\textbf{Approach 2}: When using \textit{Blue-Green deployments}, two instances of the database are available. Take the inactive database and apply the schema operations to it. In the mean time these two databases need to be kept synchronized. This can be done by using approach 1, or somehow ensure that the table under change remains in a read-only state until the active and inactive databases have switched roles.}
  \item{\textbf{Approach 3}: The last is an approach where schema operations are performed separately from the application deployments using operations which are non-blocking. As a consequence the application must be able to deal with either the old or the new database schema being in use, and construct SQL queries which can operate correctly on both schemas.}
\end{itemize}

@@

\subsubsection{Percona Toolkit~\cite{PerconaToolkit} \& Openark Kit~\cite{OpenarkKit}}
These are tools designed to help DBAs perform administrative tasks on MySQL databases. The commands \textit{oak-online-alter-table} and \textit{pt-online-schema-change} create an empty copy of the a specified table (a ghost table) and apply schema operations to the ghost table. These commands then proceed to install triggers on both tables, which insert, update, or delete the corresponding record in the other table when a record in their own table is inserted, updated, or deleted. A background task iterates over all records in the original table (in batches) and lazily copies records from the original table to the ghost table. When all records have been copied the original table and ghost table switch names atomically, ensuring that clients now use the ghost table when querying the original table name.

\subsubsection{TableMigrator~\cite{TableMigrator}}
This is a tool which works like Openark Kit and Percona Toolkit. Instead of using triggers to synchronize the two tables with each other,  TableMigrator requires tables to have an additional column which stores the date and time that record was last modified. Using this information corresponding records in both tables are synchronized with each other by passing over all records in multiple passes. Once a pass can be done relatively quickly, TableMigrator acquires an exclusive write lock on the original table --- blocking the entire table --- and proceeds to do a final pass. Once this has been done, it atomically swaps the names of the original and ghost table before releasing the write lock. Since this final pass can be done quickly, any application executing queries on that table is only very briefly blocked.

\subsubsection{SoundCloud's Large Hadron Migrator~\cite{SoundcloudLHM} \& Facebook's Online Schema Change~\cite{FacebookOSC}}
These are both tools which closely follow Approach~1 descriged earlier (Section~\ref{sec:cd}). They create a copy of a table (including its contents) and record modifications to records in the original table using triggers into a special ``deltas'' table. Once they have copied the original table and its contents, they apply the specified schema operations to the ghost table before proceeding to replay the recorded modifications from the ``deltas'' table onto the ghost table. When the ``deltas'' table is almost empty, these tools acquire an exclusive write lock on the original table --- like \textit{TableMigrator} does --- and apply the last of the recorded modifications before atomically switch the names of both tables.

@@

\subsubsection{Imago}
Imago~\cite{Dumitras:2009:WUF:1656980.1657005} is a tool which follows Approach~2. With Imago there are two production environments, called ``universes''. Imago integrates with the application and its environment at both the ingress point (load balancer) to be able to switch users from one ``universe'' to the other, and at the egress point where the (web) application sends queries to the SQL database. When a new version of the application (requiring a change to the database schema) needs to be deployed, Imago bootstraps the inactive production environment with a copy of the active database, applies the required schema operations and then copies over the data from the old universe to the new universe. When this process is almost complete, it enforces a short period of read-only access to the SQL database, while it is switching users from the old ``universe'' to the new ``universe'' using the load balancer. Imago makes no claims in regards to its support for foreign keys.

@@

\begin{itemize}
  \item None of the tools used in practice support foreign keys
  \item The tools primarily focus on MySQL databases.
  \item These tools are limited in the sense that they typically only support making small changes to the existing database schema. This forces software and release engineers to often release intermediate versions of the application, as the schema is altered in several steps.
\end{itemize}

\section{Proposed Approach} %(1.0)

@@

The previously discussed tools from practice all use Approach~1 or a variation of it. They prove that this is a viable way for simple evolutions of the database schema. However where these tools all create a single ghost table, we aim to create one ghost table for each table that is under change, and ghost tables for all tables which in turn are dependent on these tables under change. For instance adding a new non-nullable column ``summary'' to the ``books'' table (see Figure~\ref{fig:migration-stages}), means we should also create a ghost table for the ``publications`` table since there is a foreign key pointing from the ``publications'' table to the ``books'' table. The other tables do not require ghost tables, as they do not have any foreign keys referencing either the ``books'' or ``publications'' table. This approach ensures that we can support foreign keys.
